---
title: "Gene Set Enrichment Analysis using GeoMx Data"
output: html_document
date: "2026-05-01"
---

SET WORKING DIRECTORY

We first set up a working directory where I intend to store all my GeoMx files
that will be necessary to run QC and downstream analysis.

```{r WD, message=FALSE, warning=FALSE}
dir<-"/Users/srujansingh/Documents/Personal/GeoMx"
setwd(dir)
```

LOAD REQUIRED LIBRARIES

The following libraries need to be uploaded into the R working environment. These
libraries contain the functions needed to perform downstream analysis.

```{r Load Libraries, message=FALSE, warning=FALSE}
library(clusterProfiler)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(ggplot2)
library(msigdbr) #For HALLMARK Pathway
library(enrichplot) #For ClusterProfiler()
library(biomaRt)
```

LOAD THE DIFFERENTIAL EXPRESSION ANALYSIS RESULTS 

Earlier, we had performed differential expression analysis to find differentually 
expressed genes between the normal and DKD kidneys.

```{r Load results, message=FALSE, warning=FALSE}
#Loading the results from the differential expression analysis performed earlier:
results2<-readRDS("results2.rds") #results were saved as R object.
head(results2) # helps visualize the structure and elements of the result.
```

SUSBET THE RESULT


Now we subset the result to focus only on the glomerulus region. Also, we only 
keep genes with FDR<0.05

```{r Subset results, message=FALSE, warning=FALSE}
res_glom<-results2[(results2$Subset=="glomerulus")&(results2$'FDR' <0.05),]
```

ARRANGE THE GENE LIST BASED ON THE ESTIMATE VALUES

Now that we have subset the differentially expressed gene list,  we need to a
arrange the genes as per their estimate values.

```{r Arrange, message=FALSE, warning=FALSE}
df_glom<-res_glom[rev(order(res_glom$Estimate)),]
head(df_glom) #visualize after rerranging
```

CREATE THE ORDERED GENE LIST FOR GENE SET ENRICHMENT ANALYSIS

Now, we need to create the gen list with their corresponding estimate values to
run geneset enrichment analysis (GSEA). We are using using HALLMARK database for our 
geneset enrichment analysis.

```{r Create Gene List, message=FALSE, warning=FALSE}
df_glom_HALL<-df_glom$Estimate
names(df_glom_HALL)<-df_glom$Gene
head(df_glom_HALL) #visualize gene list for GSEA
```

RUN THE GENE SET ENRICHMENT ANALYSIS

Since, the ordered gene list looks fine proceeed to perform gene set enrichment 
analysis.

```{r GSEA, message=FALSE, warning=FALSE}
gene_sets <- msigdbr(species = "Homo sapiens", category = "H")
background <- gene_sets[, c("gs_name", "gene_symbol")]

#df_glom.HALL.GSEA will contain the geneset enrichment analysis results
df_glom.HALL.GSEA <- GSEA(  #GSEA function
  geneList = df_glom_HALL,
  TERM2GENE = background,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
)

```

PROCESSING THE GENE SET ENRICHMENT DATA (FOR VISUALIZATION)

```{r Process-Visualize, message=FALSE, warning=FALSE}
# df_glom.HALL.GSEA.Plot<-df_glom.HALL.GSEA
# gseaplot(df_glom.HALL.GSEA.Plot, geneSetID = 1)
# saveRDS(df_glom.HALL.GSEA.Plot, file="EnrichmentPlots/Glomerulus_Normal_vs_DKD_HALL.GSEA.rds")
# 
# if(sign(max(df_glom.HALL.GSEA$NES))==1) 
# { df_glom.HALL.GSEA<-df_glom.HALL.GSEA[rev(order(df_glom.HALL.GSEA$NES)),]
# } else { df_glom.HALL.GSEA<-df_glom.HALL.GSEA[order(df_glom.HALL.GSEA$NES),] }
# 
# df_glom.HALL.GSEA$core_enrichment<-gsub("/",",",df_glom.HALL.GSEA$core_enrichment)
# head(df_glom.HALL.GSEA$Description) #Visualize the enriched pathways.
# 
# df_glom_vis<-data.frame(Pathway=df_glom.HALL.GSEA$Description,
#                         NES=df_glom.HALL.GSEA$NES)
# 
# df_glom_vis$Pathway<-gsub("HALLMARK_","", df_glom_vis$Pathway)
# df_glom_vis$Enrichment<-"Positive"
# df_glom_vis[df_glom_vis$NES<0,]$Enrichment<-"Negative"
# head(df_glom_vis)
# 
# df_glom_vis$Pathway<-factor(df_glom_vis$Pathway, levels=rev(df_glom_vis$Pathway))

# ggplot(df_glom_vis, aes(x=Pathway, y=NES, fill=Enrichment)) + 
#   geom_bar(stat="identity") +
#   coord_flip() +
#   geom_hline(yintercept=0, linetype="solid", color="black", size=0.5) + 
#   coord_flip() + theme_minimal()+
#   scale_fill_manual(values = c("Positive" = "orangered", "Negative" = "deepskyblue")) + 
#   labs(x = "", y = "Normalized Enrichment Score", title = "Gene Set Enrichment Analysis") +
#   theme(panel.border = element_rect(color = "black", fill = NA, size = 0.5),  # Add box around plot
#         axis.text.x = element_text(size = 14, color="black"), # Increase font size of pathway names
#         axis.text.y = element_text(size = 14, color="black"),
#         axis.ticks = element_line(color = "black"),  # Add tick marks
#         axis.ticks.length = unit(0.2, "cm"),  # Adjust tick mark length
#         axis.line = element_line(color = "black"),
#         axis.title.x = element_text(size = 14))
# dev.off()

#The above code is commented out becasue it was giving error ithin the Rmd file.
#But if you run it within the R console, you should get a figure as shown below:
knitr::include_graphics("NES_GeoMx_Glom.pdf")
```

You can visualize all the enriched pathways along with their normalized enrichment scores (NES).
A positive enrichment (NES) is colored red while a negative enrichment (NES) is colored blue.

#VISUALIZE INDIVIDUAL RUNNING ENRIECHMENT SCORE PLOTS FOR INDIVIDUAL PATHWAYS

```{r Running ES, message=FALSE, warning=FALSE}
# ID1="HALLMARK_MYC_TARGETS_V1"
# gseaplot2(df_glom.HALL.GSEA.Plot, geneSetID = c(ID1),  subplots=1:2, title="MYC_TARGETS_V1_PATHWAY")

#Similarly, the above code has been commented out. But if you run it in the R console 
#you should see the follwoing plot:
knitr::include_graphics("MYC.pdf")

```

```{r Running ES2, message=FALSE, warning=FALSE}

# ID2="HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"
# gseaplot2(df_glom.HALL.GSEA.Plot, geneSetID = c(ID2),  subplots=1:2, title="EPITHELIAL_MESENCHYMAL_TRANSITION_PATHWAY")
# 
# ID3="HALLMARK_UNFOLDED_PROTEIN_RESPONSE"
# gseaplot2(df_glom.HALL.GSEA.Plot, geneSetID = c(ID3),  subplots=1:2, title="UNFOLDED_PROTEIN_RESPONSE_PATHWAY")
# 
# ID4="HALLMARK_GLYCOLYSIS"
# gseaplot2(df_glom.HALL.GSEA.Plot, geneSetID = c(ID4),  subplots=1:2, title="GLYCOLYSIS_PATHWAY")
# 
# ID5="HALLMARK_TNFA_SIGNALING_VIA_NFKB"
# gseaplot2(df_glom.HALL.GSEA.Plot, geneSetID = c(ID5),  subplots=1:2, title="TNFA_SIGNALING_VIA_NFKB_PATHWAY")

#Similarly, the above code has been commented out. But if you run it in the R console 
#you should see the follwoing plot:
knitr::include_graphics("EMT.pdf")
knitr::include_graphics("Unfolded_Protein.pdf")
knitr::include_graphics("Glycolysis.pdf")
knitr::include_graphics("NFKB.pdf")

```

Finally, you can also visualize all the running score enrichment plots within the 
same plot:

```{r Running ES_All, message=FALSE, warning=FALSE}
#gseaplot2(df_glom.HALL.GSEA.Plot, geneSetID = c(ID1, ID2, ID3, ID4, ID5),  subplots=1:2, title="ALL_ENRICHED_PATHWAYS")

#Similarly, the above code has been commented out. But if you run it in the R console 
#you should see the follwoing plot:
knitr::include_graphics("All_Pathways.pdf")
```